Contracts for unions and intersections are hard for Nickel

### 2021-05-21 paper.tex
[wip] issues in semantics

\subsection*{Code manipulation (optimizations)}
\info{Present inlining and CSE as two major code
optimizations.
Purity and immutability as two important factors.}

@@

>>>>

\subsection*{Code manipulation (optimizations)}
\label{sec:optimizations}
\info{Present inlining and CSE as two major code
optimizations.
Purity and immutability as two important factors.}

@@

Type systems featuring intersection types are hard in a fundamental way,
from their discovery, they've been studied as a way to characterize
lambda-calculus terms normalizability; which means that typechecking a
term on a system with intersection types is undecidable,
similar to how the halting problem is undecidable\todo{reference}.

Usually, when thinking about programming languages, there's an intuition
that static checks (like typechecking) are more difficult and less powerful
than dynamic checks (like testing).
For instance, it is very easy to check if,
in a given execution of a program, a variable holds an integer value; while it
may become more challenging to check if a variable will always hold an integer value,
no matter the particular execution.

As we intend to show on this paper, unions and intersections are not only
statically hard to check, but they are also dynamically hard;
\emph{they are fundamentally hard}.
In this section, we present intuitively the main challenges that arise
when attempting to develop a runtime check system for unions and intersections.

\subsubsection*{Union contracts are hard to brake}

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.
For instance, imagine having a function wrapped in a contract
that checks that maps positive numbers to positive numbers, like
function \nickel{f} on Figure \ref{fig:pos-to-pos}.

Clearly, there is an error on this program, but if the first call
to \nickel{f} that gets executed is \nickel{f 10}, the error will not
be found at that point, and the first operand of the addition would be
evaluated to \nickel{3}.
If, later on, \nickel{f 5} gets executed, the contract would get violated,
since \nickel{f} gets called with a positive value, but returns a negative
one (\nickel{-2}).

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (fun x => x - 7) @ Positive -> Positive in
(f 10) + (f 5) 
\end{lstlisting}
\caption{Wrong positive to positive function}
\label{fig:pos-to-pos}
\end{figure}

Even if the program performed two different applications of \nickel{f}
before finding the contract violation, the first one is not actually
needed, and executing only \nickel{f 5} is enough to find the problem.

Having a witness of size 1 for every contract failure is a nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated.
However, when unions get introduced, it is lost.
Consider now the program presented at Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
(f 10) + (f 5) 
\end{lstlisting}
\caption{Still wrong positive to positive or positive to non-positive function}
\label{fig:wrong-union-function}
\end{figure}

Still, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a single way to
use the wrapped value \nickel{f} that would find a violation to its contract;
at least two are needed.

This seemingly simple consequence generates major issues when considered in real
life languages.
For example, remember the different mechanisms for code manipulation presented
in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
both of these rely on the fact that substituting a pure function call by its body,
or joining equal pure expressions and sharing their result, is a semantically valid
transformation.
However, consider the examples presented at Figure \ref{fig:optimized-programs},
the first one is the inlined version of the second one, while the second
one has something like a CSE applied to it.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
(f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
let g x = f 1 x in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}
\todo{Improve example}

However, this two programs behave differently, the first one will return
\nickel{(1, "False)}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error.

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
